
# Update the MiDAS database descriptions


```{r}

# Load packages
pacman::p_load(dplyr, readxl, openxlsx, tidyverse)

# My working directory
WD = "C:/Users/HD95LP/OneDrive - Aalborg Universitet/visual_studio_code/2023_update_MIDAS_field_guide"

```


## One pipeline: 

- Runs against the current MIDAS database to catch names not previously described

# Update after new version of the MIDAS database: 

**Inputs:** 

- The downloaded webpage formatted excel file generated after last upload
- The taxonomy file of the new MIDAS version
- Change log file with the columns (tab-separted, the columns: "Before"	"After"	"Change"	"Reference (if any)"
- changes_tab_file: file with manually made changes

**Output:**

- New version of the webpage formatted excel file filtered down to only those bacteria that are updated / added
- Sheet 1: refs not updated so far


# Separate workflow (see last codeblock)
  Also possible to output all the current info about a microbe in the database
  Only requires the most recent excel from the webpage




```{r}

############################################
####             INPUTS                 ####
############################################

# Input: Newest excel (webpage format) from Marta
input_excel_file_from_Marta = paste0(WD, "/data/excel_from_Marta/2026-06-29_microbe_details.xlsx")
# Input file with manual changes:
input_change_file_manual = paste0(WD, "/data/changes_tab_file/20260629_changestab_file.xlsx") 

############################################
####           FINAL OUTPUT             ####
############################################

# Output: Final excel to upload:
output_excel_file = paste0(WD, "/output/edits_in_excel/", format(Sys.Date(), "%Y%m%d"), "", "_microbe_details_update_midas53.xlsx")

```


### Make new webpage formatted excel to upload

```{r}


source(paste0(WD, "/scripts/select_rows_and_create_excel_with_changes.R"))

make_new_excel(
  input_excel_file = input_excel_file_from_Marta,
  input_change_file = input_change_file_manual,
  output_excel_file = output_excel_file
)



```

### Check generated excel for In situ spelling and empty rows

```{r}

output_excel_file


# valid pattern
valid_regex <- "^In situ=(na|pos|neg|var)(\\[(PMID:[0-9]+|REPLACE[0-9]+)\\])*\\|\\|Other=(na|pos|neg|var)?(\\[(PMID:[0-9]+|REPLACE[0-9]+)\\])*$"

# Check for errors in reference (e.g spaces, ... )

check_sheet <- function(sheet) {
  output <- read_excel(output_excel_file, sheet = sheet)
  cols_to_test <- setdiff(colnames(output), "name")

  output %>%
    mutate(row = row_number()) %>%
    pivot_longer(all_of(cols_to_test),
                 names_to = "column",
                 values_to = "value") %>%
    mutate(valid = grepl(valid_regex, value)) %>%
    filter(!valid) %>%
    mutate(sheet = sheet) %>%     # add sheet name correctly
    select(sheet, row, name, column, value)
}


sheet_names = c("cell properties", "metabolism")
invalid_all <- map_dfr(sheet_names, check_sheet)

invalid_all %>% print(n=100)


```

### Find reference issues (spaces etc. )

```{r}

# output_excel_file = "C:/Users/HD95LP/OneDrive - Aalborg Universitet/visual_studio_code/2023_update_MIDAS_field_guide/output/edits_in_excel/20260629_microbe_details_update_midas53.xlsx"
# output <- read_excel(output_excel_file, sheet = 2)
# output <- read_excel(output_excel_file, sheet = 3)


# Sheets that should be checked
sheets_to_check <- c(
  general_information = 2,
  descriptions = 3
)

valid_reference_pattern <- "\\[(?:PMID:[0-9]+|REPLACE[0-9]+)\\]"

# Non-reference tags that are intentionally used elsewhere in the workbook.
# Adjust this list if you have additional valid MiDAS tags.
allowed_tag_names <- c(
  "DESC",
  "PATTERN",
  "NCBI",
  "PATH"
)

allowed_nonreference_pattern <- paste0(
  "\\[(?:",
  paste(allowed_tag_names, collapse = "|"),
  ")(?::[^\\]]*)?\\]"
)

# Read both sheets and convert all cells to a long table
reference_cells <- imap_dfr(
  sheets_to_check,
  function(sheet_number, sheet_name) {

    read_excel(
      output_excel_file,
      sheet = sheet_number
    ) %>%
      mutate(
        across(everything(), as.character),
        excel_row = row_number() + 1L
      ) %>%
      pivot_longer(
        cols = -c(excel_row, name),
        names_to = "column",
        values_to = "text"
      ) %>%
      transmute(
        sheet = sheet_name,
        excel_row,
        name,
        column,
        text
      ) %>%
      filter(
        !is.na(text),
        str_detect(
          text,
          regex("PMID|REPLACE", ignore_case = TRUE)
        )
      )
  }
)

# -------------------------------------------------------------------------
# 1. Complete [...] expressions that are neither valid references nor
#    explicitly permitted MiDAS metadata tags
# -------------------------------------------------------------------------

bracket_problems <- reference_cells %>%
  mutate(
    remaining_text = text %>%
      str_remove_all(valid_reference_pattern) %>%
      str_remove_all(allowed_nonreference_pattern),

    problem_fragment = str_extract_all(
      remaining_text,
      "\\[[^\\]]*\\]"
    )
  ) %>%
  select(
    sheet,
    excel_row,
    name,
    column,
    text,
    problem_fragment
  ) %>%
  unnest_longer(problem_fragment) %>%
  filter(
    !is.na(problem_fragment),
    problem_fragment != ""
  ) %>%
  mutate(
    issue = case_when(
      str_detect(
        problem_fragment,
        regex("^\\[\\s*PMID", ignore_case = TRUE)
      ) ~ "Malformed PMID tag",

      str_detect(
        problem_fragment,
        regex("^\\[\\s*REPLACE", ignore_case = TRUE)
      ) ~ "Malformed REPLACE tag",

      TRUE ~ "Unsupported square-bracket expression"
    )
  )


# -------------------------------------------------------------------------
# 2. Reference-like text outside complete square brackets
#    Examples:
#      PMID:24173461
#      [PMID:24173461
#      REPLACE52
# -------------------------------------------------------------------------

unbracketed_reference_problems <- reference_cells %>%
  mutate(
    remaining_text = text %>%
      str_remove_all(valid_reference_pattern) %>%
      str_remove_all(allowed_nonreference_pattern) %>%
      str_remove_all("\\[[^\\]]*\\]"),

    problem_fragment = str_extract_all(
      remaining_text,
      regex(
        "(?:PMID|REPLACE)[^\\s<]{0,30}",
        ignore_case = TRUE
      )
    )
  ) %>%
  select(
    sheet,
    excel_row,
    name,
    column,
    text,
    problem_fragment
  ) %>%
  unnest_longer(problem_fragment) %>%
  filter(
    !is.na(problem_fragment),
    problem_fragment != ""
  ) %>%
  mutate(
    issue = case_when(
      str_detect(
        problem_fragment,
        regex("PMID", ignore_case = TRUE)
      ) ~ "PMID reference outside valid brackets",

      str_detect(
        problem_fragment,
        regex("REPLACE", ignore_case = TRUE)
      ) ~ "REPLACE reference outside valid brackets",

      TRUE ~ "Malformed reference"
    )
  )


# -------------------------------------------------------------------------
# 3. Cells with unequal numbers of opening and closing square brackets
# -------------------------------------------------------------------------

unbalanced_bracket_problems <- reference_cells %>%
  filter(
    str_count(text, fixed("[")) !=
      str_count(text, fixed("]"))
  ) %>%
  transmute(
    sheet,
    excel_row,
    name,
    column,
    text,
    problem_fragment = NA_character_,
    issue = "Unbalanced square brackets"
  )


# -------------------------------------------------------------------------
# Combined report
# -------------------------------------------------------------------------

reference_problems <- bind_rows(
  bracket_problems,
  unbracketed_reference_problems,
  unbalanced_bracket_problems
) %>%
  distinct(
    sheet,
    excel_row,
    name,
    column,
    issue,
    problem_fragment,
    .keep_all = TRUE
  ) %>%
  select(
    sheet,
    excel_row,
    name,
    column,
    issue,
    problem_fragment,
    text
  ) %>%
  arrange(
    sheet,
    excel_row,
    column
  )

reference_problems

```



# Separate workflow: Find the old info for a bacteria 

```{r}


WD = "C:/Users/HD95LP/OneDrive - Aalborg Universitet/visual_studio_code/2023_update_MIDAS_field_guide"


# Input: Newest excel (webpage format) from Marta
  # The most resent excel file downloaded from the database
  input_excel_file_from_Marta = paste0(WD, "/data/excel_from_Marta/microbe_details_2025-08-15.xlsx")


# Output 
  # output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20240417_FISH_probes.xlsx")
  # output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20240614_Sarcinithrix.xlsx")
  # output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20241015_Turicibacter.xlsx")
  # output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20241107_proxi.xlsx")
  # output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20241107_rhoferax.xlsx")


# Read all sheets into a list of data frames
all_sheets <- lapply(excel_sheets(paste0(input_excel_file_from_Marta)), function(sheet) {
  read_excel(input_excel_file_from_Marta, sheet = sheet)
})


# Names: Define the names of which to extract
names <- tibble(name = c(
  "Nannocystis", "Polyangium", 
  "Novosphingobium", "Sphingobium", "Sphingorhabdus", "Sphingopyxis", "Sphingomonas", "Shinella",
  "Thauera", "Zoogloea", "Bradyrhizobium", "Hyphomicrobium", "Rhodobacter",
  "Haliangium", "Vulgatibacter", "Anaeromyxobacter", "Blautia", "Ca Alysiosphaera",
  "Kouleothrix", "Sphaerotilus", "Ca Amarolinea", "Ca Accumulibacter", "Leptothrix",
  "Anaerolinea", "Leptolinea", "Flexilinea", "Methanothrix", "Methanobrevibacter",
  "Nitrospira", "Nitrotoga", "Ca Microthrix", "Streptococcus", "Ca Bipolaricaulis",
  "Azonexus", "Acidovorax", "Smithella", "Methanolinea", "Methanosarcina",
  "Tetrasphaera", "Thermomonas", "Thermovirga", "Micropruina", "Microlunatus",
  "Trichococcus", "Ca Proximibacter", "Paracoccus", "Methanothermobacter", "Ferribacterium",
  "Geobacter", "Defluviicoccus", "Propionivibrio", "Ca Brevefilum", "Geothrix",
  "Ca Phosphoribacter", "Azoarcus", "Rhodoferax", "Methanoculleus", "Comamonas",
  "Propioniciclava", "Haliscomenobacter", "Sedimentibacter", "Ca Villigracilis", "Ca Competibacter",
  "Ca Opimibacter", "OLB13", "Ca Brachybacter", "Ca Parvibacillus", "Nitrosomonas", "Nitrosospira", 
  "Gordonia", "Turicibacter", "Pelotomaculum", "Skermania", "Ca Epiflobacter", "Herbinix", "Ca Cloacimonas", "Neomegalonema", "P3OB-42", "KD-3-10"
)
)

names <- tibble(name = c(
"Ca Epiflobacter"))

distinct(names)
length(names$name)

# Function:  Get all the info from each sheet in the excel
index = c(1:5)
find_old_info_make_long_format <- function(sheet_index = 2) {
  
  exist <- all_sheets[[sheet_index]] %>% 
    #filter(name %in% unique(names$name)) %>%    # If this line is removed all bacteria are included
    pivot_longer(cols = -name, names_to = "column_name_for_change", values_to = "new_value") 
  
  return(exist)
}

# Apply the function: 
take_old_info <- 
  rbind(
    find_old_info_make_long_format(sheet_index = 2), 
    find_old_info_make_long_format(sheet_index = 3), 
    find_old_info_make_long_format(sheet_index = 4),
    find_old_info_make_long_format(sheet_index = 5)
    ) %>% 
  #filter(new_value != "In situ=na||Other=na") %>% 
  filter(column_name_for_change == "FISH probes") %>%    # If this is included eg. all fish probes are extracted
  #filter(column_name_for_change == "Description") %>% 
  #arrange(new_value)
  arrange(name)

###############################################################################
### Manual version - save one excel 
###############################################################################
  
output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20250911_PHN_meeting/20250911_FISH_probes.xlsx")

  output_excel_file <- "NAME_YOUR_EXCEL"
  # Create a new Excel workbook
  output_wb <- createWorkbook()
  # Add a worksheet to the new Excel workbook
  addWorksheet(output_wb, sheetName = "change_file")
  # Write the filtered data to the new worksheet
  writeData(output_wb, x = take_old_info, sheet = "change_file")
  openxlsx::saveWorkbook(output_wb, output_excel_file)


###############################################################################
### Automatic version - Save each taxa as seperate file
###############################################################################

take_old_info_apply_version <- function(tax = names$name[1]){

  # subset to only one taxa
  take_old_info_sub <- 
    take_old_info %>% 
    filter(name == tax)
  # Create a new Excel workbook
  output_wb <- createWorkbook()
  # Add a worksheet to the new Excel workbook
  addWorksheet(output_wb, sheetName = "change_file")
  # Write the filtered data to the new worksheet
  writeData(output_wb, x = take_old_info_sub, sheet = "change_file")
  # Generate the output Excel file path
  output_excel_file <- 
    paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20250911_PHN_meeting/",   # <- change the output folder 
           format(Sys.Date(), "%Y%m%d"), "_", 
           str_replace(tax, " ", "_"), ".xlsx")
  # Save the workbook
  openxlsx::saveWorkbook(output_wb, output_excel_file)
}

lapply(names$name, take_old_info_apply_version)


```






