# Update the MiDAS database descriptions

```{r}

# Load packages
pacman::p_load(dplyr, readxl, openxlsx, tidyverse)


WD = "C:/Users/HD95LP/OneDrive - Aalborg Universitet/visual_studio_code/2023_update_MIDAS_field_guide"





```


## One pipeline: 

- Runs against the current MIDAS database to catch names not prevoisly described

# Update after new version of the MIDAS database: 

**Inputs:** 

- The downloaded webpage formatted excel file generated after last upload
- The taxomony file of the new MIDAS version
- Change log file with the columns (tab-separted, the columns: "Before"	"After"	"Change"	"Reference (if any)"
- changes_tab_file: file with manually made changes

**Output:**

- New version of the webpage formatted excel file filtered down to only thos bacteria that are updated / added
- Sheet 1: refs not updated so far


```{r}

############################################
####             INPUTS                 ####
############################################

# Input: Newest excel (webpage format) from Marta
input_excel_file_from_Marta = paste0(WD, "/data/excel_from_Marta/20240614_microbe_details_v2.xlsx")
# Input file with manual changes:
input_change_file_manual = paste0(WD, "/data/changes_tab_file/20240614_changes_v2.xlsx") 

# Input: Change log with new names
input_MIDAS_change_log_file = paste0(WD, "/data/MIDAS_change_log/MIDAS_changelog_53.txt")
# Input: Newest version Midas database 
midas_db_tax_file = paste0(WD, "/../data_microbial_groups/MiDAS/MIDAS_53_taxa.txt")



############################################
####           INTERMEDIATES            ####
############################################

# Output: Excel with new names  
output_excel_file_all_names = paste0(WD, "/output/include_all_names/MIDAS_53_excel_web_format.xlsx")


# Output (intermediate); changes tab file with the info from old names 
output_change_file_old_info = paste0(WD, "/output/changes_tab_file/20240301_MIDASv52_data_from_old_tax_names.xlsx")  

# Output: File with merged manual changes (manual + those from old names) 
output_merged_changes_tab = paste0(WD, "/output/merged_changes_tab_file/20240301_merged_changes_tab_file_v2.xlsx")


############################################
####           FINAL OUTPUT             ####
############################################

# Output: Final excel to upload:
output_excel_file = paste0(WD, "/output/edits_in_excel/", format(Sys.Date(), "%Y%m%d"), "_microbe_details_update_midas53.xlsx")

```



### Update the webpage formatted excelfile with all new names

Do you need to add descriptions for new genera or species without any info in the database? 

  - Could be in case of a new database version
  - Add description of a species 
    
If not: Skip this    

```{r}

source(paste0(WD, "/scripts/get_all_new_names.R"))

missing_names(WD = WD, 
              # Input: Newest excel (webpage format) from Marta
              input_excel_file = input_excel_file_from_Marta,
              # Input:Newest version Midas database 
              midas_db = midas_db_tax_file, 
              # Output: Excel with new names  
              output_excel_file = output_excel_file_all_names
              )

```


### Get info from old names transfered to the new names

Name change: Do you need to add old information to a new taxonomy?

  - Could be in case of a new database version
  - Name update 
    
If not: Skip this  


```{r}

source(paste0(WD, "/scripts/update_names_but_retain_old_data.R"))

get_old_info(
    WD = WD, 
    # Input: Excel file WITH ALL names from new database
    input_excel_file = output_excel_file_all_names,
    # Input: Change log with new names
    input_MIDAS_change_log = input_MIDAS_change_log_file, 
    # Output: Tab changes file
    output_excel_file = output_change_file_old_info, old_midas_version = "MiDAS 5.1"
    )

### Load new changes with manual created changes_tab_file and merge with the old info

change_file_old_info <- read_excel(output_change_file_old_info) %>% 
  rename("value_from_old_name" = "new_value")
change_file_new_manual_info <- read_excel(input_change_file_manual) 

# Merged change files and when values are dublicated keep only new 
merged_change_file <- full_join(change_file_old_info, change_file_new_manual_info) %>% 
  mutate(new_value = ifelse(is.na(new_value), value_from_old_name, new_value)) %>% 
  distinct(genus_name, column_name_for_change, new_value) 

# Save to new merged change file: 

  ##Create a new Excel workbook
  output_wb <- createWorkbook()
  # Add a worksheet to the new Excel workbook
  addWorksheet(output_wb, sheetName = "Sheet1")
  # Write the filtered data to the new worksheet
  writeData(output_wb, sheet = "Sheet1", merged_change_file)

# Save the new Excel workbook
saveWorkbook(output_wb, output_merged_changes_tab)

```


### Make new webpage formatted excel to upload

```{r}

## If NO old was taken - thus no reason to merge change files : 
input_change_file_x = input_change_file_manual
## If old WAS USED - thus merge change files : 
input_change_file_x = output_merged_changes_tab


# No new names 
output_excel_file_all_names <- input_excel_file_from_Marta


source(paste0(WD, "/scripts/select_rows_and_create_excel_with_changes.R"))

make_new_excel(
  input_excel_file = output_excel_file_all_names,
  input_change_file = input_change_file_x,
  output_excel_file = output_excel_file
)



```


#### Find the old info for a group of bacteria 

```{r}

WD = "C:/Users/HD95LP/OneDrive - Aalborg Universitet/visual_studio_code/2023_update_MIDAS_field_guide"


# Input: Newest excel (webpage format) from Marta

# Output 
output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20240417_FISH_probes.xlsx")
output_excel_file <- paste0(WD, "/output/get_old_data_to_use_for_manual_changes/20240614_Propionivibrio.xlsx")



input_excel_file_from_Marta = paste0(WD, "/data/excel_from_Marta/20240614_microbe_details_v2.xlsx")


# Read all sheets into a list of data frames
all_sheets <- lapply(excel_sheets(paste0(input_excel_file_from_Marta)), function(sheet) {
  read_excel(input_excel_file_from_Marta, sheet = sheet)
})


# Names 
# names <- read_excel(paste0(WD, "/data/functional_data/20240207_functional groups_MiDAS5.2.xlsx")) %>% 
#   rename("name" = 1)
names <- 
  #tibble(name = c("Acidovorax", "Giesbergeria"))
  tibble(name = c("Propionivibrio"))



# Function 
index = c(1:5)
find_old_info_make_long_format <- function(sheet_index = 2) {
  
  exist <- all_sheets[[sheet_index]] %>% 
    filter(name %in% unique(names$name)) %>%  
    pivot_longer(cols = -name, names_to = "column_name_for_change", values_to = "new_value") 
  # %>% 
  #   filter(!is.na(new_value))
  #   
  return(exist)
}

take_old_info <- 
  rbind(
    find_old_info_make_long_format(sheet_index = 2), 
    find_old_info_make_long_format(sheet_index = 3), 
    find_old_info_make_long_format(sheet_index = 4),
    find_old_info_make_long_format(sheet_index = 5)
    ) %>% 
  #filter(new_value != "In situ=na||Other=na") %>% 
  #filter(column_name_for_change == "FISH probes") %>%
  #filter(column_name_for_change == "Description") %>% 
  arrange(name)

# Create a new Excel workbook
output_wb <- createWorkbook()

# Add a worksheet to the new Excel workbook
addWorksheet(output_wb, sheetName = "change_file")

# Write the filtered data to the new worksheet
writeData(output_wb, x = take_old_info, sheet = "change_file")

openxlsx::saveWorkbook(output_wb, output_excel_file)




```






